---
title: "Data Visualization Strategies in R: Part 2"
author: "Justin Joque (Clark Data Labs)"
date: "12/12/2016" 
output: 
  html_document:
      toc: true # table of content true
      toc_float: 
         collapsed: false
         smooth_scroll: false
      depth: 3  # display up to three depths of headings (specified by #, ## and ###)
      number_sections: false  # if you want number sections at each table header
      theme: united  # one of many options for theme
   
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval=FALSE, tidy = TRUE, fig.align='center')
```

# Tutorial Outline

This tutorial is the second of a two-part series providing some basic instruction on how to (a) visualize data, and (b) use R for visualizations.

**Part 2** Includes: 

1. **Visualizing 'Complext' Data (3+ variables)**
2. **More Types of Visualizations**
3. **Some Statistics**

*To see* **Part 1**, *visit: INSERT LINK HERE*

# 1. Visualizing 'Complex' Data (3+ variables)

```{r}
#multiple scatterplots can be made by choosing multiple variables (this will plot columns 5 through 11)
plot(census[5:11], cex=.7)

#tying other information to variables; here we take the unemployment and use it as the size
plot(census$rent,census$hh_income,cex=census$unemploy*18,pch=20,col="red") #pch=20 chooses a different symbol

#we can also build up a plot by using the points function i.e.
points(census$rent,census$hh_income,pch=20,col="blue")
```


# 2. More Types of Visualizations


```{r}
x <- census$area #assign area data (from census) to variable named x  
y <- census$population #assign population data (from census) to variable named y
labelfield <- census$name #assign name data (from census) to variable named labelfield
plot(x,y) #graph y v. x
identify(x,y,labels=labelfield) #label points interactively #click on points to be labeled, press 'esc' key to see results
```


# 3. Some Statistics
```{r}
#sample statistics
############

#t-test
t.test(census$inc_male_15_w,census$inc_female_15_w)
#paired t-test
t.test(census$inc_male_15_w,census$inc_female_15_w,paired=TRUE)

#correlation
cor.test(census$bach_degree,census$unemploy)

#linear models
#############

plot (census$unemploy ~ census$bach_degree)
model <- lm(census$unemploy ~ census$bach_degree)
plot(model)

```

