---
title: "datavis_r_script"
author: "M.C.M. Simao"
date: "12/12/2016" 
html_document: default
pdf_document: default
md_document:
    variant: markdown_github
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval=FALSE, tidy = TRUE, fig.align='center')
```

# Data Visualization Strategies in R

## Introduction

### Outline
* Quick Tips on R
* Understanding your Data
* Organizing/Manipulating Data
* Choosing Appropriate Visualization(s)
* Visualizing 'Simple' Data (1-2 variables)
* Visualizing 'Complex' Data (3+ variables)
* More Types of Visualizations
* Some Statistics

## Quick Tips on R
**INSERT INFO ON COMMENTS AND ? AND PACKAGES? LINKs LIKE MAPPING_R?**
```{r}
#a few helpful tricks
########
c(1,2,3) #concatanate a set of elements
1:20 #list numbers between two numbers
colors() #list all of R's colors
palette() #show the current order of colors
palette(rainbow(6)) #set colors
palette(c("green","brown","red")) #set user defined colors
identify(x,y,labels=labelfield) #label points interactively
```


## Understanding your Data
```{r}
#loading the data

#we can call the data frame anything we want, but I've used census for this example
census <- read.csv(file.choose()) #read in the csv file
head(census) #see the headers plus the first six rows
summary(census) #see some summary statistics
fix(census) #open an excel-esque data editor
```

## Organizing/Manipulating Data

```{r}
#use this formula in excel or R to map a variable to be between two numbers (i.e. mapping a variable to use for size between 1 and 10):
#(new_max - new_min) * ([value] - lowest_value) / (highest_value - lowest_value) + new_min

#it is easy to add a new column:
#census$newcol <- [formula]

#subsetting data
################

# census[1:5] returns the first 5 columns, as does census[,1:5], but census[1:5,] give the first five rows

upper <- census[census$region=="upper",] #note the comma after the conditional!
central <- census[census$region=="central",] #without the comma it defaults to the second part
lower <- census[census$region=="south",]

```


### Scaling

### Normalizing
```{r}
#we can plot the log of a variable by simply addiing it to the plot command
plot(census$hh_income, log(census$population,2))

#for the y axis this is the population normalized by area (population density)
plot(census$hh_income, census$population/census$area)
```

### Aggregating


## Choosing Appropriate Visualization(s)

## Visualizing 'Simple' Data (1-2 variables)
### 1 Variable
```{r}
#histogram
############

hist(census$unemploy,xlab="Unemployment",main="Histogram of Unemployment by County in Michigan") 
#by adding breaks as a paramter we can increase or decrease the bin-size

#boxplot
###########
boxplot(census[,12],census[,13],col="lightblue",names=c("men","women"),main="Median Income by Sex For Population\nThat Worked Full-time Last 12 Months\nby County ACS 2006-2010")
boxplot(hh_income ~ region,data=census, col="lightgreen", main="Median Income by Region\nby County ACS 2006-2010") #this will give us household in come by region

```
### 2 Variables
```{r}
#scatter plot
##############

plot(census$unemploy,census$hh_income)
#cex controls the size of the points (1 is default)

#subsetting data (repeat from Organizing/Manipulating Data Above)
################

# census[1:5] returns the first 5 columns, as does census[,1:5], but census[1:5,] give the first five rows

upper <- census[census$region=="upper",] #note the comma after the conditional!
central <- census[census$region=="central",] #without the comma it defaults to the second part
lower <- census[census$region=="south",]

#we can also build up a plot by using the points function i.e.
plot(c(0,max(census$rent)),c(0,max(census$hh_income)),type="n") #create a blank canvas
points(upper$rent,upper$hh_income,pch=20,col="blue")
points(central$rent,central$hh_income,pch=20,col="red")
points(lower$rent,lower$hh_income,pch=20,col="green")
```

## Visualizing 'Complex' Data (3+ variables)

```{r}
#multiple scatterplots can be made by choosing multiple variables (this will plot columns 5 through 11)
plot(census[5:11], cex=.7)

#tying other information to variables; here we take the unemployment and use it as the size
plot(census$rent,census$hh_income,cex=census$unemploy*18,pch=20,col="red") #pch=20 chooses a different symbol

#we can also build up a plot by using the points function i.e.
points(census$rent,census$hh_income,pch=20,col="blue")
```


## More Types of Visualizations

## Some Statistics
```{r}
#sample statistics
############

#t-test
t.test(census$inc_male_15_w,census$inc_female_15_w)
#paired t-test
t.test(census$inc_male_15_w,census$inc_female_15_w,paired=TRUE)

#correlation
cor.test(census$bach_degree,census$unemploy)

#linear models
#############

plot (census$unemploy ~ census$bach_degree)
model <- lm(census$unemploy ~ census$bach_degree)
plot(model)

```

